---
title: "Special topics"
author: "Carson Sievert & Joe Cheng"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    self_contained: false
vignette: >
  %\VignetteIndexEntry{3. Special topics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  fig.align = "center",
  out.width = "80%",
  class.output = "R",
  comment = ""
)
options(shiny.suppressMissingContextError = TRUE)
```

## Meta counterparts to `observeEvent()` & `eventReactive()`

**shinymeta** currently does not provide meta-counterparts for `eventReactive()` and `observeEvent()`, but it's possible to 'hand-roll' your own counterpart from existing building blocks (namely, `metaReactive2()`, `metaObserve2()`, and `metaExpr()`) Both of these functions are essentially a wrapper for a common reactive pattern where you want `isolate()` all reactive values except for one value/input. For example, if you want to

```{r, eval = FALSE}
r <- eventReactive(input$x, {
  c(input$x, input$y)
})
```

is equivalent to:

```{r, eval = FALSE}
r <- reactive({
  req(input$x)
  isolate({
    c(input$x, input$y)
  })
})
```

so, to create the meta-counterpart:

```{r, eval = FALSE}
r <- metaReactive2({
  req(input$x)
  isolate(metaExpr({
    c(input$x, input$y)
  }))
})
```

Similarly, for `observeEvent()`:

```{r, eval = FALSE}
observeEvent(input$x, {
  message(input$x + input$y)
})
```

is equivalent to:

```{r, eval = FALSE}
observe({
  req(input$x)
  isolate({
    message(input$x, input$y)
  })
})
```

so, to create the meta-counterpart:

```{r, eval = FALSE}
o <- metaObserve2({
  req(input$x)
  isolate(metaExpr({
    message(input$x + input$y)
  }))
})
```

**Warning:** Do not attempt to use existing `eventReactive()`/`observeEvent()` by calling `metaExpr()` within their handler bodies. It won't work with either one. `eventReactive()` won't work because it caches its results, oblivious to whether it's in normal mode or meta mode; and `observeEvent()` won't work because (non-meta) observers don't even have a way to return values, period.


## Shiny modules

**shinymeta** is designed to work with Shiny modules, [here's an example](https://github.com/rstudio/shinymeta/blob/main/inst/examples/modules/app.R).
Modifying an existing Shiny app that uses modules may be more involved than simply capturing domain logic and marking reactive reads. For example, if you have one or more `callModule()` calls which create a (namespaced) output object(s), you may want to have that module function return the output object so you can `expandChain()` various meta-outputs from different modules at the same time ([as done in the example app](https://github.com/rstudio/shinymeta/blob/main/inst/examples/modules/app.R#L32)).

## Shiny, tidyeval, and shinymeta

> TL;DR: The same steps in the [overview](#overview) will work for a Shiny app that uses [tidyeval](https://tidyeval.tidyverse.org/). When the reactive read is unquoted directly with `!!`, the generated code is the same as what you'd write by hand. When the conversion is bound to an intermediate variable, or when you're splicing multiple values with `!!!`, the generated code is functionally correct but reads slightly differently from hand-written code. Switching to `varSelectInput()` or to a function variant that takes character strings (e.g. `across(all_of(...))`) is the simplest fix where possible.

Most **tidyverse** functions evaluate code expressions in a special context (e.g., they search for names within a data frame). That's how **dplyr** knows, for example, to lookup names (e.g. `cyl`) and evaluate calls (e.g., `mean(mpg)`) within a context defined by `mtcars`:

```{r}
library(dplyr)

# compute mean miles per gallon (mpg) by cylinder (cyl)
mtcars %>%
  group_by(cyl) %>%
  summarise(avg = mean(mpg))
```

This approach makes for an expressive interactive interface, but it also complicates things if we wish to pass variables into (i.e., program around) these functions (because they quote their arguments). For example, if you had a variable, named `var`, that represented another name with the column name of interest, **dplyr** thinks you're looking for a column named `var`, not `mpg`:

```{r}
var <- as.name("mpg")

mtcars %>%
  group_by(cyl) %>%
  summarise(avg = mean(var))
```

To workaround this problem, **tidyverse** functions allow you to unquote (i.e., replace a name with it's value) via the `!!` operator. Just to demonstrate, if we unquote `var`, we'd get back the name (i.e., symbol) `mpg`.

```{r}
rlang::expr(!!var)
```

That's why this code gives us the desired result of average miles per gallon (`mpg`) per cylinder (`cyl`).

```{r}
mtcars %>%
  group_by(cyl) %>%
  summarise(avg = mean(!!var))
```

Often times in a Shiny app we wish to pass an input value to a **tidyverse** function argument (as a variable). When the input is a character string, that requires converting it into a symbolic name with `as.symbol()` or `rlang::sym()` and then unquoting the result with `!!`. For example, here's a Shiny app to compute the mean of different `mtcars` variables by cylinder (`cyl`).

```{r, eval = FALSE}
library(shiny)
library(tidyverse)
ui <- fluidPage(
  selectInput("var", "Select a variable", names(mtcars)),
  verbatimTextOutput("out")
)
server <- function(input, output) {
  output$out <- renderPrint({
    mtcars %>%
      group_by(cyl) %>%
      summarise(mean_mpg = mean(!!sym(input$var)))
  })
}
shinyApp(ui, server)
```

Adding **shinymeta** support in this case is straight-forward. As with any other app, you'll have to capture the domain logic (i.e., wrap `renderPrint()` with `metaRender()`), then mark reactive read `..()`.

```{r, eval = FALSE}
server <- function(input, output) {
  output$out <- metaRender(renderPrint, {
    mtcars %>%
      group_by(cyl) %>%
      summarise(mean_mpg = mean(!!sym(..(input$var))))
  })
  observe(print(expandChain(output$out())))
}
```

In version v1.2.0, **shiny** introduced `varSelectInput()` essentially to remove the need to convert character string(s) into symbolic name(s). For example, in the app below, `input$var` already represents the symbolic name of interest, so you can do:

```{r, eval = FALSE}
ui <- fluidPage(
  varSelectInput("var", "Select a variable", mtcars),
  verbatimTextOutput("out")
)
server <- function(input, output) {
  output$out <- renderPrint({
    mtcars %>%
      group_by(cyl) %>%
      summarise(mean_mpg = mean(!!input$var))
  })
}
shinyApp(ui, server)
```

As in the other examples, you can mark the reactive read with `..()` (before unquoting with `!!`) and the code generation should "just work". Technically speaking, this works because, when `..()` encounters a symbolic name that it doesn't recognize, it returns the code *to generate* the symbol instead of the bare symbol (i.e., it returns `as.symbol("mpg")` instead of `mpg` which makes the `!!` work in both normal and meta execution). When deparsed, `metaExpr()` recognizes the `!!as.symbol("foo")` pattern and collapses it back to the bare symbol `foo`, so the generated code matches what you'd write by hand.

```{r, eval = FALSE}
server <- function(input, output) {
  output$out <- metaRender(renderPrint, {
    mtcars %>%
      group_by(cyl) %>%
      summarise(mean_mpg = mean(!!..(input$var)))
  })
  observe(print(expandChain(output$out())))
}
shinyApp(ui, server)
```

The examples above generate code identical to what a human would write. Two patterns still produce code that differs slightly from hand-written form.

The first is the intermediate-variable form of the `sym()` conversion — binding the result of `sym(..(input$var))` to a name and then unquoting that name with `!!`:

```{r, eval = FALSE}
server <- function(input, output) {
  output$out <- metaRender(renderPrint, {
    var_sym <- sym(..(input$var))
    mtcars %>%
      group_by(cyl) %>%
      summarise(mean_mpg = mean(!!var_sym))
  })
}
shinyApp(ui, server)
```

The generated code preserves both the `var_sym` assignment and the `!!var_sym` reference. The fix is the inline form `mean(!!sym(..(input$var)))` shown earlier in this section.

The second is multi-value splicing with `!!!`:

```{r, eval = FALSE}
ui <- fluidPage(
  selectInput("var", "Select variables", names(mtcars), multiple = TRUE),
  verbatimTextOutput("out")
)
server <- function(input, output) {
  output$out <- metaRender(renderPrint, {
    select(mtcars, !!!syms(..(input$var)))
  })
  observe(print(expandChain(output$out())))
}
shinyApp(ui, server)
```

When `input$var` comes from `selectInput(multiple = TRUE)`, the generated code preserves the `!!!syms(...)` call rather than expanding it inline. The code is correct and runs as expected, but `!!!syms(c("mpg", "cyl"))` isn't what you'd write by hand. The simplest workaround is to use a function that takes character strings directly. For column selection specifically, the tidyselect helper `all_of()` works directly inside `select()`:

```{r, eval = FALSE}
server <- function(input, output) {
  output$out <- metaRender(renderPrint, {
    select(mtcars, all_of(..(input$var)))
  })
  observe(print(expandChain(output$out())))
}
shinyApp(ui, server)
```
